PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-06-21 manuscript_00.Rmd
rejigger

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
c_d <- plot_grid(c,d, rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.6))
```

@@

## Cell Type Proportions
```{r, fig.width=5, fig.height=12}

o_ct <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
p_data <- meta_filter %>% 
  #mutate(CellType_predict = Tissue) %>% 
  #mutate(CellType_predict = case_when(!CellType_predict %in% ws_oct ~ 'Other',
  #                                    TRUE ~ CellType_predict)) %>% 
  filter(Organ == 'Eye') %>% 
  mutate(Citation = str_trunc(Citation, width = 20)) %>% 
  group_by(Citation, organism, Tissue, CellType_predict) %>% 
  summarise(Count = n()) %>% 
  mutate(Proportion = Count/sum(Count) * 100)

p_data %>% ggplot(aes(y=Citation,x=Proportion,fill=CellType_predict)) + geom_bar(stat= 'identity') +
  scale_fill_manual(name= meta_column,
                    values = color_list) + 
  ggforce::facet_col(vars(organism, Tissue), scale = 'free', space = 'free') +
  cowplot::theme_cowplot() +
  guides(fill=guide_legend(ncol =1))
```
## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types

@@

# number of celltype (predict) - organism - study counts (with >=50 cells)
meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% 
  pivot_wider(values_from = n, names_from = organism) %>% 
  filter(!is.na(CellType_predict)) %>% 
  replace(is.na(.), values = 0) %>% 
  arrange(CellType_predict) %>% 
  rename(`CellType (Predict)` = CellType_predict) %>% 
  mutate(`Well Supported` = case_when(`Homo sapiens` > 2 & `Mus musculus` > 2 ~ 'Yes',
                                      TRUE ~ 'No')) %>% 
  flextable()
```

# Well supported ocular cell types identify

@@

```{r}
ws_oct <- meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% filter(n >= 3) %>% 
  ungroup() %>% 
  group_by(CellType_predict) %>% 
  summarise(Count = n()) %>% 
  filter(Count > 1, !is.na(CellType_predict)) %>% 
  pull(CellType_predict)

@@

```{r, fig.height=8, fig.width=6}
g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=5, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)

@@

base_yo <- consist_diff %>% filter(Symbol %in% genes) %>% pull(Base)

hm <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
                              column_split = cts, 
                              row_split = base_yo, 
                              row_labels = genes, 
                              column_title_rot = 90, 
                              name = 'log2FC', use_raster = TRUE,
                              row_title_rot = 0)
ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 30, 2), "mm"))
```
```{r}
# top 100 genes in each organism/base category
diff_tab %>%

  group_by(Base, Organism) %>% 
  top_n(., 100, -padj) %>% 
  arrange(Base, Organism) %>% 
  relocate(Base, Organism) %>% 
  select(-Against, -Group)
```

>>>>

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(glue)
#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
data_dir_vGiga <- '~/data/scEiaD_gigascience/'
data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
c_d <- plot_grid(c,d, rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))

# make E now

o_ct <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
p_data <- meta_filter %>% 
  #mutate(CellType_predict = Tissue) %>% 
  #mutate(CellType_predict = case_when(!CellType_predict %in% ws_oct ~ 'Other',
  #                                    TRUE ~ CellType_predict)) %>% 
  filter(Organ == 'Eye') %>% 
  mutate(Citation = paste(str_trunc(Citation, width = 20), str_trunc(batch, side = "left", width = 15))) %>% 
  group_by(Citation, organism, Tissue, CellType_predict) %>% 
  summarise(Count = n()) %>% 
  mutate(Proportion = Count/sum(Count) * 100)

proportion_bar_plot <- p_data %>% ggplot(aes(y=Citation,x=Proportion,fill=CellType_predict)) + geom_bar(stat= 'identity') +
  scale_fill_manual(name= meta_column,
                    values = color_list) + 
  ggforce::facet_col(vars(organism, Tissue), scale = 'free', space = 'free') +
  cowplot::theme_cowplot() +
  guides(fill=guide_legend(ncol =1))

plot_grid(a_aa_b, c_d, proportion_bar_plot, ncol = 3, rel_widths = c(1,1.6, 3))
```

@@

## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types

@@

# number of celltype (predict) - organism - study counts (with >=50 cells)
ws_table <- meta_filter %>%  
  filter(Organ == 'Eye') %>% 
  group_by(organism, CellType_predict, study_accession) %>% 
  count() %>% 
  filter(n >= 50) %>% 
  ungroup() %>% 
  group_by(organism, CellType_predict) %>% 
  count() %>% 
  pivot_wider(values_from = n, names_from = organism) %>% 
  filter(!is.na(CellType_predict)) %>% 
  replace(is.na(.), values = 0) %>% 
  arrange(CellType_predict) %>% 
  rename(`CellType (Predict)` = CellType_predict) %>% 
  mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
                                      TRUE ~ 'No'))
ws_table %>% 
  flextable()
```

## Well supported ocular cell types identify

@@

```{r}
ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)

@@

```{r, fig.height=9, fig.width=7}
g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=5, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)

@@

base_yo <- genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)

hm <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
                              column_split = cts, 
                              row_split = base_yo, show_row_dend = FALSE, 
                              row_labels = genes, 
                              column_title_rot = 90, 
                              name = 'log2FC', use_raster = TRUE,
                              row_title_rot = 0)
ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 30, 2), "mm"))
```

## Tease apart closely related cell types
```{r}
x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Base %in% c("Early RPC", "Late RPC", "RPC")) %>%  
  filter(Group == 'CellType (Predict)', 
         Against %in% c("Early RPC", "Late RPC", "RPC")) %>% 
  group_by(Base, Against, Gene) %>% 
  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-3)) %>% 
  arrange(-`mean log2FC`) %>%   
  collect() %>% 
  filter(!is.na(`mean log2FC`))

  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
  left_join(gene_tab, by = 'GENEID') %>% 
  relocate(SYMBOL, GENEID) %>% 
  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
  arrange(-`mean log2FC`)
```
